### 2014-12-16 text/Chapter04.tex
Spell checking

The development in this chapter can be seen as a counterpart to the well-known development of 
\emph{effect systems} \cite{effects-gifford} and the use of \emph{monads} \cite{monad-notions}
in programming languages. The syntax and type system of the flat coeffect calculus follows 
asimilar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure, as explained in the previous chapter, most importantly in lambda abstraction
(the relationship with monads is further discussed in Section~\ref{sec:flat-related}).

Wadler and Thiemann \cite{monads-effects-marriage} famously showed a correspondence between effect 
systems to monads (a topic that has been also discussed by Atkey \cite{monads-parameterised-notions}
and recently revisited by Vazou an Leijen \cite{monads-effects-remarriage}). This line of work
relates effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
$\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar correspondence between 
\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
defining the semantics in terms of comonadic computations is not a simple mechanical dualization
of the work on effect systems and monads.

@@

The flat coeffect calculus is defined in terms of \emph{flat coeffect algebra}, which defines
the structure of context annotations, such as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
sets of implicit parameters, versions represented as integers or other values. The expressions of 
the calculus are those of the $\lambda$-calculus with \emph{let} binding; assuming $\alpha$ ranges 
over base types, the types of the calculus are defined as:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \alpha \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a 
context $r$.

@@

Because of the assymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
the duality between monads and comonads can no longer help us with defining the additional structure
required to model full $\lambda$-calculus. In comonadic computations, additional information is attached
to the context. In application and lambda abstraction, the context is propagated differently than 
in effectful computations.

@@

\noindent
The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combineds them into a 
single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful 
as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.

@@

The (\emph{internalized sequencing}) rule is useful when defining a concrete semantics for 
a language that provides constructs for explicitly providing the context. For example, consider
a language with implicit parameters where a parameter is defined by $\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2$.
Semantics for such language would provide a reduction rule:
%
\begin{equation*}
\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~(\kvd{glet}~x = \ident{?p}~\kvd{in}~e)
 ~~\leadsto~~ \kvd{let}~x = e_1~\kvd{in}~e
\end{equation*}
%
The \kvd{glet} construct provides a way of sequentialising the context-require\-ments so
that they can be discharged by matching constructs providing the required context. As discussed
tfearlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
this form of evaluation (rather than looking at semantics for concrete systems).

@@

\begin{theorem}[Type preservation for call-by-value]
\label{thm:cbv-reduction}
In a flat coeffect system with the (\emph{approxi\-mation}) property, 
if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
Consider the typing derivation for the term $(\lambda x.e)\,v$ before reduction:
\begin{equation*}
\inference
  { \inference
  { \inference
      {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash e : \tau_2}
      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 } &
   \coctx{\Gamma}{\cunit} \vdash v : \tau_1 }
  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{t})} \vdash (\lambda x.e)\,v : \tau_2} }
  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; \cclrd{t}} \vdash (\lambda x.e)\,v : \tau_2}
\end{equation*}
The final step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
%
\begin{equation*}
\tyrule{sub}
 { \coctx{\Gamma}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
 { \coctx{\Gamma}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
\end{equation*}
%
Comparing the final conclusions of the above two typing derivations shows that 
the reduction preserves type and coeffect annotation.
\end{proof}

@@

Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
actually access a previous value of the argument $x$. This cannot be caputred by a flat coeffect 
system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.

@@

\begin{lemma}[Top-pointed substitution]
\label{thm:cbn-substitution-top}
In a top-pointed flat coeffect calculus with an aglebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
not change the coeffects of $e_r$:
%
\begin{equation*}
\begin{array}{l}
 \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
   \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
 \quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau_r
\end{array}
\end{equation*}
\end{lemma}

@@

\paragraph{Semantics.} We follow the same approach as in Section~\ref{sec:flat-semantics} and use
categorical semantics to explain (and confirm) the design of the sub-typing rules. The semantics
of a judgement $\sem{\tau <: \tau'}$ is a function $\sem{\tau} \rightarrow \sem{\tau'}$. 
As shown in Figure~\ref{fig:flat-semantics-sub}, the seamntics of the sub-typing rule (\emph{typ}) 
then just composes the semantics of the original expression with the conversion produced by the 
semantics of the sub-typing judgement.

@@

\begin{remark}
\label{thm:flat-alt-abs}
For a flat coeffect calculus with a strictly oriented algebra, the standard (abs) rule can 
be derived from the (idfun) rule.
\end{remark}
\begin{proof}
The following derives the conclusion of (\emph{abs}) using (\emph{idabs}), sub-coeffecting, 
sub-typing and the fact that the algebra is \emph{stricly oriented}:

@@

\begin{theorem}
\label{thm:flat-tup-eq}
For a flat coeffect calculus with pairs and units, the following holds:
%
\begin{equation*}
\begin{array}{rclcl}
 \ident{assoc}~(e_1, (e_2, e_3)) &\equiv& ((e_1, e_2), e_3) &\qquad\qquad&(\emph{associativity}) \\
 \pi_1~(e, ()) &\equiv& e &&(\emph{right unit})\\
 \pi_2~((), e) &\equiv& e &&(\emph{left unit})\\
\end{array} 
\end{equation*}
\end{theorem}
\begin{proof}
Follows from the fact that $(\C, \cpar, \czero)$ is a monoid and \ident{assoc}, $\pi_1$ and
$\pi_2$ are pure functions (treated as contstants in the langauge).
\end{proof}

@@

Implicit parameters can be captured by a monad, but \emph{just} a monad is not enough.
Lambda abstraction in effect systems does not provide a way of splitting the context
requirements between declaration site and call site (or, semantically, combining the implicit 
parameters available in the scope where the function is deined and those specified by the caller).

\paragraph{Categorical relationship.}
Before looking at the necessary extensions, consider the two ways of modelling implicit 
parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
are:
%
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
 \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
\end{array} 
\end{equation*}
%
The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the 
two representations are equivalent:
%
\begin{remark}
Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
reader monad via currying/uncurrying isomoprhism.
\end{remark}
\begin{proof}
The isomorphism is demonstrated by the following equation:
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 &\narrow{=}&
 (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
 &\narrow{=}& \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) =
 \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
\end{array}
\end{equation*}
\end{proof}

@@

\paragraph{Delaying effects in monads.} 
In the syntax of the language, the above difference is manifested by the (\emph{abs}) rules for 
monadic effect systems and comonadic coeffect systems. The following listing shows the two rules 
side-by-side, using the effect system notation for both of them:
%
\begin{equation*}
\tyrule{cabs}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s} }
  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 \;\&\; \cclrd{r}}
\quad
\tyrule{mabs}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s}}
  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 \;\&\; \cclrd{\emptyset} }
\qquad
\end{equation*}
%
In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
the monadic rule (\emph{mabs}) places all requirements on the call site. This follows from the
fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
%
\begin{equation*}
\sem{\lambda x.e} \;=\; \ident{unit}~(\lambda x. \sem{e})
\end{equation*}
%
The type of $\ident{unit}$ is $\alpha\rightarrow\mtyp{\alpha}{\cclrd{\emptyset}}$, but in this specific
case, the $\alpha$ is instantiated to be $\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}$ and so this
use of \ident{unit} has a type:
%
\begin{equation*}
\ident{unit}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{\emptyset}}(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2})
\end{equation*}
%
In order to split the implicit parameters of the body ($\cclrd{r \cup s}$ on the left-hand side) between 
the declaration site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
call site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
(which we call \ident{delay}) with the following signature:
%
\begin{equation*}
\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{r}}(\tau_1\rightarrow\mtyp{\cclrd{s}}{\tau_2})
\end{equation*}
%
The operation reveals the difference between effects and coeffects -- intuitively, given a function
with effects $\cclrd{r \cup s}$, it should execute the effects $\cclrd{r}$ when wrapping the 
function, \emph{before} the function actually perforsm the effectful operation with the effects.
The remaining effects $\cclrd{s}$ are delayed as usual, while effects $\cclrd{r}$ are removed
from the effect annotation of the body.

@@

This chapter presented the \emph{flat coeffect calculus} -- a unified system for tracking 
\emph{whole-context} properties of computations, that is properties related to the 
execution environment or the enitre context in which programs are executed.
This is the first of the three \emph{coeffect calculi} developed in this thesis.

>>>>

The development in this chapter can be seen as a counterpart to the well-known development of 
\emph{effect systems} \cite{effects-gifford} and the use of \emph{monads} \cite{monad-notions}
in programming languages. The syntax and type system of the flat coeffect calculus follows 
a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure, as explained in the previous chapter, most importantly in lambda abstraction
(the relationship with monads is further discussed in Section~\ref{sec:flat-related}).

Wadler and Thiemann \cite{monads-effects-marriage} famously showed a correspondence between effect 
systems to monads (a topic that has been also discussed by Atkey \cite{monads-parameterised-notions}
and recently revisited by Vazou an Leijen \cite{monads-effects-remarriage}). This line of work
relates effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
$\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar correspondence between 
\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
defining the semantics in terms of comonadic computations is not a simple mechanical dualisation
of the work on effect systems and monads.

@@

The flat coeffect calculus is defined in terms of \emph{flat coeffect algebra}, which defines
the structure of context annotations, such as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
sets of implicit parameters, versions represented as integers or other values. The expressions of 
the calculus are those of the $\lambda$-calculus wwith \emph{let} binding; assuming $\alpha$ ranges 
over base types, the types of the calculus are defined as:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \alpha \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a 
context $r$.

@@

Because of the asymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
the duality between monads and comonads can no longer help us with defining the additional structure
required to model full $\lambda$-calculus. In comonadic computations, additional information is attached
to the context. In application and lambda abstraction, the context is propagated differently than 
in effectful computations.

@@

\noindent
The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combines them into a 
single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful 
as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.

@@

The (\emph{internalized sequencing}) rule is useful when defining a concrete semantics for 
a language that provides constructs for explicitly providing the context. For example, consider
a language with implicit parameters where a parameter is defined by $\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2$.
Semantics for such language would provide a reduction rule:
%
\begin{equation*}
\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~(\kvd{glet}~x = \ident{?p}~\kvd{in}~e)
 ~~\leadsto~~ \kvd{let}~x = e_1~\kvd{in}~e
\end{equation*}
%
The \kvd{glet} construct provides a way of sequentialising the context-require\-ments so
that they can be discharged by matching constructs providing the required context. As discussed
earlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
this form of evaluation (rather than looking at semantics for concrete systems).

@@

\begin{theorem}[Type preservation for call-by-value]
\label{thm:cbv-reduction}
In a flat coeffect system with the (\emph{approxi\-mation}) property, 
if $\ctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
Consider the typing derivation for the term $(\lambda x.e)\,v$ before reduction:
\begin{equation*}
\inference
  { \inference
  { \inference
      {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash e : \tau_2}
      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 } &
   \coctx{\Gamma}{\cunit} \vdash v : \tau_1 }
  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{t})} \vdash (\lambda x.e)\,v : \tau_2} }
  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; \cclrd{t}} \vdash (\lambda x.e)\,v : \tau_2}
\end{equation*}
The final step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
%
\begin{equation*}
\tyrule{sub}
 { \coctx{\Gamma}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
 { \coctx{\Gamma}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
\end{equation*}
%
Comparing the final conclusions of the above two typing derivations shows that 
the reduction preserves type and coeffect annotation.
\end{proof}

@@

Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.

@@

\begin{lemma}[Top-pointed substitution]
\label{thm:cbn-substitution-top}
In a top-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
not change the coeffects of $e_r$:
%
\begin{equation*}
\begin{array}{l}
 \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
   \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
 \quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau_r
\end{array}
\end{equation*}
\end{lemma}

@@

\paragraph{Semantics.} We follow the same approach as in Section~\ref{sec:flat-semantics} and use
categorical semantics to explain (and confirm) the design of the sub-typing rules. The semantics
of a judgement $\sem{\tau <: \tau'}$ is a function $\sem{\tau} \rightarrow \sem{\tau'}$. 
As shown in Figure~\ref{fig:flat-semantics-sub}, the semantics of the sub-typing rule (\emph{typ}) 
then just composes the semantics of the original expression with the conversion produced by the 
semantics of the sub-typing judgement.

@@

\begin{remark}
\label{thm:flat-alt-abs}
For a flat coeffect calculus with a strictly oriented algebra, the standard (abs) rule can 
be derived from the (idfun) rule.
\end{remark}
\begin{proof}
The following derives the conclusion of (\emph{abs}) using (\emph{idabs}), sub-coeffecting, 
sub-typing and the fact that the algebra is \emph{strictly oriented}:

@@

\begin{theorem}
\label{thm:flat-tup-eq}
For a flat coeffect calculus with pairs and units, the following holds:
%
\begin{equation*}
\begin{array}{rclcl}
 \ident{assoc}~(e_1, (e_2, e_3)) &\equiv& ((e_1, e_2), e_3) &\qquad\qquad&(\emph{associativity}) \\
 \pi_1~(e, ()) &\equiv& e &&(\emph{right unit})\\
 \pi_2~((), e) &\equiv& e &&(\emph{left unit})\\
\end{array} 
\end{equation*}
\end{theorem}
\begin{proof}
Follows from the fact that $(\C, \cpar, \czero)$ is a monoid and \ident{assoc}, $\pi_1$ and
$\pi_2$ are pure functions (treated as constants in the language).
\end{proof}

@@

Implicit parameters can be captured by a monad, but \emph{just} a monad is not enough.
Lambda abstraction in effect systems does not provide a way of splitting the context
requirements between declaration site and call site (or, semantically, combining the implicit 
parameters available in the scope where the function is defined and those specified by the caller).

\paragraph{Categorical relationship.}
Before looking at the necessary extensions, consider the two ways of modelling implicit 
parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
are:
%
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
 \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{eader monad})
\end{array} 
\end{equation*}
%
The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the 
two representations are equivalent:
%
\begin{remark}
Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
reader monad via currying/uncurrying isomorphism.
\end{remark}
\begin{proof}
The isomorphism is demonstrated by the following equation:
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 &\narrow{=}&
 (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
 &\narrow{=}& \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) =
 \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
\end{array}
\end{equation*}
\end{proof}

@@

\paragraph{Delaying effects in monads.} 
In the syntax of the language, the above difference is manifested by the (\emph{abs}) rules for 
monadic effect systems and comonadic coeffect systems. The following listing shows the two rules 
side-by-side, using the effect system notation for both of them:
%
\begin{equation*}
\tyrule{cabs}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s} }
  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 \;\&\; \cclrd{r}}
\quad
\tyrule{mabs}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s}}
  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 \;\&\; \cclrd{\emptyset} }
\qquad
\end{equation*}
%
In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
the monadic rule (\emph{mabs}) places all requirements on the call site. This follows from the
fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
%
\begin{equation*}
\sem{\lambda x.e} \;=\; \ident{unit}~(\lambda x. \sem{e})
\end{equation*}
%
The type of $\ident{unit}$ is $\alpha\rightarrow\mtyp{\alpha}{\cclrd{\emptyset}}$, but in this specific
case, the $\alpha$ is instantiated to be $\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}$ and so this
use of \ident{unit} has a type:
%
\begin{equation*}
\ident{unit}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{\emptyset}}(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2})
\end{equation*}
%
In order to split the implicit parameters of the body ($\cclrd{r \cup s}$ on the left-hand side) between 
the declaration site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
call site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
(which we call \ident{delay}) with the following signature:
%
\begin{equation*}
\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{r}}(\tau_1\rightarrow\mtyp{\cclrd{s}}{\tau_2})
\end{equation*}
%
The operation reveals the difference between effects and coeffects -- intuitively, given a function
with effects $\cclrd{r \cup s}$, it should execute the effects $\cclrd{r}$ when wrapping the 
function, \emph{before} the function actually performs the effectful operation with the effects.
The remaining effects $\cclrd{s}$ are delayed as usual, while effects $\cclrd{r}$ are removed
from the effect annotation of the body.

@@

This chapter presented the \emph{flat coeffect calculus} -- a unified system for tracking 
\emph{whole-context} properties of computations, that is properties related to the 
execution environment or the entire context in which programs are executed.
This is the first of the three \emph{coeffect calculi} developed in this thesis.
